---
title: "GDP and Life Expectancy Across Continents"
author: "Balpreet Kaur Saluja (24202918)"
date: today
format: pdf
execute:
  echo: false
---

## Introduction

The **gapminder** dataset, available in the `gapminder` R package, contains historical data on GDP per capita, life expectancy, and population for countries around the world. This document explores the relationship between GDP per capita and life expectancy in *2007*, faceted by continent.

## Descriptive Statistics

Below is a brief summary of the main variables used in the plot:

- *GDP per capita*: Measures the economic output per person.
- *Life expectancy*: Average number of years a person is expected to live.
- The data is from the year **2007**, the most recent available year in the dataset.

## GDP vs Life Expectancy

We use a scatter plot to visualize the relationship between **GDP per capita** and **life expectancy**, colored by continent.

::: {.text-info}
Each point represents a country. The plot reveals a positive correlation — countries with higher GDP per capita generally have higher life expectancy.
:::


```{r, warning=FALSE, message=FALSE}
#Loading required libraries
library(ggplot2)
library(gapminder)
library(dplyr)

#Filtering the data for 2007
data_2007 <- gapminder %>% filter(year == 2007)

#Creating the scatter plot
ggplot(data_2007, aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_x_log10() + 
  labs(
    title = "GDP per Capita vs Life Expectancy (2007)",
    x = "GDP per Capita (log scale)",
    y = "Life Expectancy (years)",
    color = "Continent"
  ) +
  theme_minimal()
```

```{r}
# Calculating mean life expectancy by continent
avg_lifeExp <- data_2007 %>%
  group_by(continent) %>%
  summarise(mean_lifeExp = mean(lifeExp))

# Creating the bar plot
ggplot(avg_lifeExp, aes(x = continent, y = mean_lifeExp, fill = continent)) +
  geom_col(width = 0.6) +
  labs(
    title = "Average Life Expectancy by Continent (2007)",
    x = "Continent",
    y = "Average Life Expectancy (years)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## Conclusion

This visual shows how socioeconomic factors such as GDP per capita are often linked to public health indicators like life expectancy. Africa tends to cluster lower on both metrics, while Europe shows high values in both.
